\begin{abstract}
\emergencystretch=2em
In multi-agent social settings, behavior and model reliability vary significantly across relationships. Beyond accurately inferring what other agents will do, an agent must also calibrate how confidently to act on those inferences---and do so independently for each relationship. An agent may carry a well-validated model of one partner, a fragile model of another, and an actively revising model of a third; collapsing these into a single confidence estimate loses information that matters for action. Affect has traditionally been described in terms of valence and arousal, but it can also serve a functional role: regulating how strongly beliefs are expressed in action rather than shaping their content. Using the active inference framework, we model affect as a metacognitive estimate of confidence in one's current partner model, and formalize this by extending affective precision to be relationship-specific. Each partner's behavioral evidence updates its own confidence estimate, which modulates how strongly the agent commits to its inferred partner state during action selection.

Simulations in a multi-partner trust game show that partner-local affective precision tracks predictive reliability rather than realized value: a reliably cooperative and a reliably defecting partner can both support high precision, because what they share is predictability, not positive value. In open decision regimes, behavioral effects arise through action sharpening rather than improved partner-state inference. Confidence supports decisive action when the partner model remains valid, but can impair adaptation when a partner changes abruptly, indicating that affective precision is a calibration mechanism rather than a guaranteed payoff-improvement mechanism. Varying precision gain and prior model fitness reveals a family of trust-calibration profiles, providing a computational basis for individual differences in social trust.

Together, these results describe affective precision as a relationship-specific metacognitive signal that separates social prediction from social commitment.

\keywords{Active inference \and Affective precision \and Social metacognition \and Multi-agent systems \and Trust \and Individual differences}
\end{abstract}
