\section{Introduction}

Social decision-making often unfolds across relationships that differ in reliability. In repeated cooperative exchange, the same agent may interact with a familiar collaborator, an uncertain newcomer, and a partner whose behavior is predictable but adversarial. What matters is therefore not only whether a partner is valuable, but whether the agent has a reliable model of how that partner tends to behave. Repeated social exchange depends on learning such partner-specific regularities, including reciprocity, reputation, moral character, and social value~\cite{KingCasas2005Trust,Behrens2008SocialLearning,Delgado2005MoralCharacter}. In these settings, the question is not only what an agent believes about each partner, but how confidently those beliefs should guide policy inference.

This separates social decision-making into three related computational problems. The first is partner inference: learning what a partner is like, including whether their behavior is cooperative, exploitative, reciprocal, or unstable. The second is theory of mind: modeling what another agent knows, intends, or expects, and how those representations guide their actions~\cite{FrithFrith2012,Hampton2008Mentalizing,Pitliya2025ToMActiveInference}. The third is confidence calibration: estimating whether the focal agent's current partner model is reliable enough to guide policy commitment. Accurate partner inference and sophisticated mentalizing do not by themselves solve this third problem. Knowing that someone is predictable is not the same as knowing whether they are desirable; knowing that someone is unreliable is not the same as knowing how cautiously to act toward them. An agent that infers accurately but commits poorly wastes its social knowledge, while one that commits confidently on fragile models exposes itself to exploitation.

We model social affect as a metacognitive estimate of whether the focal agent's current partner model warrants confident action. This treats affect not as a direct readout of another agent's intentions, nor as a record of reward and punishment, but as a regulator of how strongly current beliefs are expressed in policy. In psychology and neuroscience, affect is often described in terms of valence and arousal---the felt pull toward or away from situations and social commitments~\cite{Barrett2006EmotionParadox}. Functionally, however, affect can also assign action-relevant weight to what an agent already infers, shaping the degree of commitment rather than the content of inference.

The active inference framework~\cite{Friston2010FreeEnergy,Friston2017ProcessTheory,Parr2022ActiveInference} provides a formal route from this observation to a computational mechanism. In active inference, policy inference is precision-weighted: the same beliefs can produce decisive or tentative behavior depending on \(\gamma\), the parameter controlling how sharply an agent commits to its inferred policy distribution. Prior active-inference accounts of affect and emotion link affective dynamics to model evidence, prediction error, emotional-state inference, and precision-weighted policy inference~\cite{joffily2013emotional,smith2019simulating,Hesp2021DeeplyFeltAffect,pattisapu2024free}. Social-learning work similarly shows that agents revise expectations from discrepancies between expected and observed behavior of others, including in trust-game interactions~\cite{joiner2017social}. However, in multi-partner settings, the evidence relevant to confidence often comes from particular partners rather than from the social environment as a whole. We therefore define partner-local affective precision from partner-response likelihood surprisal: the surprise of an observed partner response under the focal agent's current posterior over that partner's type and stance.

This mechanism complements, rather than replaces, existing accounts of social inference. Theory-of-mind models in active inference address how agents model what others know or intend through recursive belief structures and factorized generative models~\cite{Pitliya2025ToMActiveInference,RuizSerra2025Factorised}. Volatility and social-learning models address how prediction errors should change belief updating under uncertainty~\cite{Mathys2011HGF,Behrens2008SocialLearning,Diaconescu2017Hierarchical}. Partner-local affective precision instead estimates how strongly each current partner model should be expressed in policy. Each partner receives its own \(\beta_k\) estimate, updated only from that partner's behavioral evidence, which modulates policy precision when inferring policies toward that partner.

We evaluate this mechanism in a repeated multi-partner graded trust game with partner-choice conditions, abrupt betrayal, and profile-style parameter variation. The simulations test whether behavioral effects arise through policy sharpening rather than improved partner-state inference, whether partner-local precision reshapes engagement under choice and change, and whether precision gain together with prior model fitness shapes temporally structured profiles of social trust revision.

The remainder of the paper is organized as follows. Section~2 specifies the multi-partner trust game, per-partner generative models, and the partner-local affective precision mechanism. Section~3 reports simulation results. Section~4 discusses implications, limitations, and directions for future work.
