\section{Discussion}
\label{sec:discussion}

The results establish partner-local affective precision as a calibration mechanism for social policy commitment. Across simulations, precision followed partner-response predictability more closely than realized payoff, lost its behavioral effect when the $\beta_k \rightarrow \gamma_k$ pathway was cut, and changed deployment under abrupt partner change without guaranteeing reward improvement. These findings support the central separation of the paper: social prediction and social commitment are related but distinct problems. An agent may know what a partner is likely to do while still needing to estimate how confidently that model should guide policy.

This separation explains why payoff is not the primary criterion for the mechanism. A reliably cooperative partner and a reliably defecting partner can both generate low partner-response likelihood surprisal because both are predictable, even though their payoff consequences differ. Conversely, a favorable but inconsistent partner can remain difficult to model. Partner-local affective precision therefore does not encode partner value directly. It estimates how well the current partner model is working, then regulates how strongly that model is expressed in policy. As a computational hypothesis, this predicts that social familiarity or confidence may follow predictability rather than positive value, so agents may reengage predictable-but-harmful partners or remain uncertain around favorable-but-inconsistent partners.

The tracked-only ablation gives a direct check on the mechanism: when $q(\beta_k)$ was updated but prevented from modulating $\gamma_k$, behavior matched the no-affect baseline. Thus, the behavioral effect does not come from improved partner-state inference; it appears when partner-local confidence reaches policy selection and changes how sharply the agent commits to its current partner model.

This mechanism complements existing accounts of social inference by targeting a different part of the control loop. Active-inference accounts of affect and emotion motivate links between model fit, precision, and action~\cite{joffily2013emotional,smith2019simulating,Hesp2021DeeplyFeltAffect,pattisapu2024free}. Theory-of-mind models specify richer beliefs about what another agent knows or intends~\cite{Pitliya2025ToMActiveInference,RuizSerra2025Factorised}, while volatility and social-learning models regulate how strongly prediction errors should revise beliefs under uncertainty~\cite{Mathys2011HGF,Behrens2008SocialLearning,Diaconescu2017Hierarchical}. Partner-local affective precision instead regulates how strongly the focal agent should express each current partner model in policy. A richer social agent could combine these components: infer what partners believe, estimate confidence in each partner model, detect when that confidence has become stale, and redirect exploration or commitment accordingly.

The betrayal and profile analyses show why this extra change-detection layer matters. Partner-local precision is most useful when accumulated confidence still reflects a valid relationship. After abrupt change, the same confidence can remain behaviorally active even though the evidence that produced it came from an earlier partner state. The mechanism therefore reshapes deployment reliably, but does not by itself decide when old confidence should be discounted. Future work should add volatility or change-detection mechanisms that monitor shifts in prediction-error structure and reduce stale confidence when a previously predictable partner begins behaving unexpectedly.

The profile experiments should be interpreted as computational hypotheses about trust calibration rather than fitted psychological categories. Motivated by work on social learning, psychosis, and attachment~\cite{Behrens2008SocialLearning,Adams2013ComputationalPsychosis,Cittern2018AttachmentActiveInference}, they show how precision gain and prior model fitness jointly shape update speed, confidence amplitude, sensitivity to change, reengagement after repair, and post-betrayal commitment errors. High gain expands confidence swings without monotonic payoff improvement; low gain stabilizes behavior but slows revision; cautious priors protect against premature confidence while suppressing confidence even toward reliable partners. These profiles therefore generate predictions about social trust calibration, not claims about stable human types without behavioral fitting.

Several extensions would make the mechanism more complete. First, $\beta_k$ is currently an auxiliary tracker outside the generative model; embedding it as a hidden state would allow the agent to form expectations about its own future policy precision. Second, partner-choice policies could become explicitly exploratory after detected change, allowing the agent to sample alternatives rather than only adjusting commitment toward the current partner. Third, environment-side partners could be replaced with reciprocal active-inference agents that maintain their own beliefs, policies, and precision estimates.

The most direct empirical extension is to fit the model to trust-game time series. The simulations produce measurable quantities such as exploitation sensitivity, betrayal recovery time, reengagement latency after repair, and post-betrayal commitment errors. Fitting $\alpha$ and prior model-fitness parameters to individual participants would test whether the proposed profiles correspond to stable differences in social trust calibration. The simulation targets that should be validated first are the paper's core contrasts: predictability over value, deployment over inference, confidence under abrupt change, and the non-monotonic effects of gain and prior model fitness.

\section{Conclusion}
\label{sec:conclusion}

This paper formalized social affect as a relationship-specific metacognitive signal: an estimate of how confidently an agent should act on its current model of each partner. Implemented as partner-local affective precision, this signal updates from partner-response likelihood surprisal and modulates policy precision during selection. Across simulations, it separated social prediction from social commitment: precision tracked predictability more closely than payoff, affected behavior through policy deployment rather than improved partner-state inference, and remained sensitive to the temporal structure of trust revision.

The resulting picture is not that social confidence should simply be maximized. Precision gain and prior model fitness create different calibration profiles, each with tradeoffs in confidence accumulation, betrayal adaptation, reengagement, and repair. Volatility detection, adaptive confidence discounting, and reciprocal partner models would extend the mechanism, but the core target is now clear: to test whether human trust-game behavior shows the same separations between predictability and value, inference and deployment, and reengagement and restored confidence. Partner-local affective precision offers a computational account of social metacognition in which relationships differ not only in what an agent believes about them, but in how strongly those beliefs should guide action.
