\section{Affective Precision Update and Policy Selection}
\label{app:precision_update}

This appendix gives the $\beta$-update rule, policy-precision mapping, and partner-choice policy-selection procedure.

\subsection{Categorical beta support}

Each partner maintains a categorical posterior $q(\beta_k)$ over discrete inverse-precision levels $\{0.5, 0.67, 1.0, 1.5, 2.0\}$, updated each round from affective charge $\phi_k$. The categorical representation keeps the inverse-precision convention explicit, bounds the precision dynamics, and allows the update to be analyzed as movement over interpretable confidence states.

\subsection{Persistence prior}

The update proceeds in three steps. First, a persistence prior smooths the previous posterior via a tridiagonal transition matrix:
\begin{equation}
p(\beta_k) \leftarrow T \cdot q_{\text{prev}}(\beta_k),
\label{eq:persistence}
\end{equation}
where $T$ is tridiagonal with persistence parameter 0.8 and reflecting boundaries, encoding the prior expectation that precision evolves slowly.

\subsection{Charge-based pseudo-likelihood}

Second, a charge-based pseudo-likelihood maps positive charge toward high policy precision (low $\beta$) and negative charge toward low policy precision (high $\beta$):
\begin{equation}
\log \ell(\beta_\ell) = \phi_k \cdot \frac{1}{\beta_\ell},
\label{eq:pseudo_lik}
\end{equation}
where $\beta_\ell$ ranges over the five support points. This update satisfies the intended monotonicity: positive charge reinforces lower $\beta$, negative charge reinforces higher $\beta$, and the effect scales with effective precision at each support point.

\subsection{Posterior update and policy precision}

Third, the posterior is normalized:
\begin{equation}
q(\beta_k) \propto \exp(\log \ell(\beta_\ell)) \cdot p(\beta_k).
\label{eq:posterior_update}
\end{equation}
Policy precision is then set from the posterior mean as given in Eq.~\eqref{eq:gamma_update} of the main text. The quantity $\beta_k$ is inverse precision, so higher posterior mean $\beta_k$ lowers policy precision $\gamma_k$, while lower posterior mean $\beta_k$ raises $\gamma_k$. The $\beta_k$ posterior is maintained as an external auxiliary tracker rather than a proper hidden state in the generative model, which simplifies implementation at the cost of losing the agent's ability to form prior expectations over its own future affective states (discussed in Section~\ref{sec:discussion}).

\subsection{Cross-partner policy selection}

In partner-choice regimes, partner-specific precision sharpens or flattens policy commitment within a relationship while preserving the mean cross-partner evidence for engaging that partner. The transformation in Eq.~\eqref{eq:centered_policy_scores} applies $\gamma_k$ to deviations from the partner-specific policy mean, rather than to raw policy scores. The agent samples or selects policies from the combined set of transformed scores across all partners and policies. This lets $\gamma_k$ control commitment among policies involving partner $k$ without changing the average evidence for approaching partner $k$ during cross-partner comparison.

\subsection{Ablation variants}

The four affect-runtime variants are defined in Section~2.5. For completeness: the no-affect control disables the $\beta$ tracker entirely and fixes $\gamma_k = \gamma_{\text{base}}$; the tracked-only ablation updates $q(\beta_k)$ normally but fixes $\gamma_k = \gamma_{\text{base}}$, allowing the precision estimate to be analyzed without influencing policy selection; the default partner-local variant applies the full update pipeline described in B.1--B.4 and sets $\gamma_k$ from each partner's posterior mean; and the shared-$\beta$ ablation maintains a single pooled $\beta$ state updated from all partner evidence while keeping per-partner POMDP beliefs intact, with all partner branches inheriting precision from the shared expected $\beta$.
