\section{Focal-Agent POMDP Specification}
\label{app:pomdp}

This appendix specifies the focal agent's partner-local generative model in standard discrete active-inference form. For each partner \(k\), the agent maintains a POMDP \(\mathcal{M}_k\) with hidden factors for partner type, partner stance, and own-investment bookkeeping. The model is fit to the generative process in Appendix~\ref{app:generative_process}, but it is not identical to the environment: payoff is represented as an observation modality whose likelihood marginalizes over partner response, scheduled betrayal and repair are environment-side interventions, and affective precision is maintained outside the POMDP.

\subsection{Partner-local factorization}

The focal agent maintains a separate model \(\mathcal{M}_k\) for each partner \(k\), rather than one pooled social model. Each \(\mathcal{M}_k\) contains its own observation likelihoods, transition dynamics, preferences, initial priors, and policy priors. This factorization keeps partner-specific evidence local to the relationship that generated it. Partner choice is handled outside the individual partner-local POMDPs by comparing candidate partner--investment policies after each partner model has produced policy evidence.

\subsection{Hidden state factors and observations}

As described in Section~\ref{sec:partner_models}, the focal agent's generative model for each partner $k$ maintains three hidden factors:
\begin{align}
s_k^{\text{type}} &\in \mathcal{S}^{\text{type}}, \label{eq:s_type} \\
s_k^{\text{stance}} &\in \mathcal{S}^{\text{stance}}, \label{eq:s_stance} \\
s^{\mathrm{own}} &\in \mathcal{A},
\end{align}
where
\begin{align*}
\mathcal{S}^{\text{type}} &= \{\text{cooperator, reciprocator,} \\
  &\quad \text{exploiter, random}\}, \\
\mathcal{S}^{\text{stance}} &= \{\text{trusting, neutral, hostile}\}, \\
\mathcal{A} &= \{0,1,2,3,4,5\}.
\end{align*}
Executed investment levels index \(s^{\mathrm{own}}\). The own-investment factor is deterministic bookkeeping: the agent always knows its executed investment level, and this factor makes that knowledge available to the observation model. It is not a psychological hidden variable. The joint hidden state per partner spans \(4\times3\times6=72\) configurations.

Observations are partner response and focal payoff:
\begin{align}
o_k^{\text{act}} &\in \mathcal{O}^{\text{act}}, \\
o_k^{\text{pay}} &\in \mathcal{O}^{\text{pay}},
\end{align}
where $\mathcal{O}^{\text{act}}=\{\text{cooperate, defect}\}$ and $\mathcal{O}^{\text{pay}}$ comprises eleven discrete payoff levels (Eq.~\eqref{eq:graded_payoff_map}).

\subsection{Observation likelihoods}
\label{app:pomdp_likelihoods}

\paragraph{Partner-response likelihood \(A^{\mathrm{act}}_k\).}
The partner-response modality represents the focal agent's likelihood model for observed cooperate-or-defect responses. It uses the cooperation probabilities in Appendix~\ref{app:process_response}: \(P(o^{\mathrm{act}}_k=\mathrm{cooperate}\mid s^{\mathrm{type}}_k,s^{\mathrm{stance}}_k)\), with \(P(o^{\mathrm{act}}_k=\mathrm{defect}\mid \cdot)=1-P(o^{\mathrm{act}}_k=\mathrm{cooperate}\mid \cdot)\).
The modality is uniform over \(s^{\mathrm{own}}\): partner response depends on inferred partner type and stance, not on the bookkeeping factor itself. Investment affects future response probabilities indirectly through stance transitions.

\paragraph{Payoff likelihood \(A^{\mathrm{pay}}_k\).}
In the environment, payoff is deterministic given investment and partner response (Appendix~\ref{app:process_payoff}). In the focal agent's POMDP, payoff is represented as a likelihood over payoff observations conditional on own investment and latent partner state, marginalizing over partner response:
\begin{align}
&P(o^{\mathrm{pay}}_k=v \mid s^{\mathrm{own}},s^{\mathrm{type}}_k,s^{\mathrm{stance}}_k) \nonumber\\
&\quad =
\sum_{o^{\mathrm{act}}\in\{\mathrm{cooperate},\mathrm{defect}\}}
\mathbb{I}\!\left[\pi(s^{\mathrm{own}},o^{\mathrm{act}})=v\right]
P(o^{\mathrm{act}}_k=o^{\mathrm{act}} \mid s^{\mathrm{type}}_k,s^{\mathrm{stance}}_k).
\label{eq:a_pay_marginal}
\end{align}
This makes higher investment beneficial when cooperation is likely and costly when defection is likely, while keeping payoff preferences inside the generative model.

\subsection{Transition dynamics}

\paragraph{Type drift \(B_k^{\mathrm{type}}\).}
Partner type is uncontrollable and drifts slowly:
\begin{equation}
P(s_k^{\mathrm{type}\prime}=j\mid s_k^{\mathrm{type}}=i)=
\begin{cases}
1-p_{\mathrm{switch}} & j=i,\\
p_{\mathrm{switch}}/(K_{\mathrm{type}}-1) & j\neq i,
\end{cases}
\label{eq:b_type}
\end{equation}
with $K_{\mathrm{type}}=4$ and default $p_{\mathrm{switch}}=0.05$. Scheduled betrayal scenarios set $p_{\mathrm{switch}}=0$.

\paragraph{Stance dynamics \(B_k^{\mathrm{stance}}\).}
The focal agent's stance-transition model uses the same investment-dependent interpolation form as the environment-side process in Appendix~\ref{app:process_stance}: \(\rho(a)=a/(|\mathcal{A}|-1)=a/5\) and \(B^{\mathrm{stance}}(a)=\rho(a)B^{\mathrm{high}}+(1-\rho(a))B^{\mathrm{low}}\).
The endpoint matrices \(B^{\mathrm{high}}\) and \(B^{\mathrm{low}}\) are given in Appendix~\ref{app:process_stance}. This transition model encodes the agent's prior belief that higher investment tends to build trust, lower investment or withholding tends to damage trust, and recovery from hostility is slower than collapse.

Scheduled betrayal and repair protocols intervene on the environment-side stance process. The focal agent is not given explicit prior knowledge of these intervention times; it must adapt from subsequent partner responses and payoff observations.

\paragraph{Own-investment bookkeeping \(B^{\mathrm{own}}\).}
Identity transition on the executed investment level (not an inferred belief):
\begin{equation}
P(s^{\mathrm{own}\prime}=a^\prime \mid s^{\mathrm{own}}, a)=\mathbb{I}[a^\prime=a].
\label{eq:b_own}
\end{equation}

\subsection{Preferences, priors, and policy priors}

\paragraph{Observation preferences \(C_k\).}
There is no direct preference over partner responses:
\begin{equation}
C_k^{\mathrm{act}} = [0,\,0] \quad \text{over } \{\mathrm{cooperate},\mathrm{defect}\}.
\label{eq:c_act}
\end{equation}
Payoff preferences ascend over the graded payoff support generated by Eq.~\eqref{eq:graded_payoff_map}:
\begin{equation}
C_k^{\mathrm{pay}} = \mathrm{log\_softmax}\!\left(\frac{[5,6,7,8,9,10,10.5,11,11.5,12,12.5]}{\tau}\right),
\label{eq:c_pay}
\end{equation}
where $\tau$ is a temperature parameter controlling preference sharpness. Instrumental motivation therefore enters through payoff observations and multi-step planning rather than direct preferences over partner responses.

\paragraph{Initial-state priors \(D_k\).}
Type, stance, and own-investment priors are
\begin{align}
D_k^{\mathrm{type}} &= \left[\tfrac{1}{4},\tfrac{1}{4},\tfrac{1}{4},\tfrac{1}{4}\right], \label{eq:d_type}\\
D_k^{\mathrm{stance}} &= [0.2,\,0.6,\,0.2], \label{eq:d_stance}\\
D^{\mathrm{own}} &= \mathrm{Unif}(\{0,1,2,3,4,5\}). \label{eq:d_own}
\end{align}

\paragraph{Policy priors \(E_k\).}
Policy priors are uniform over admissible graded investment sequences at the configured planning horizon. In partner-choice settings, partner selection is handled outside the per-partner POMDP by evaluating partner--investment policy candidates with precision-adjusted scores before executing the chosen investment. This cross-partner comparison is specified in Appendix~\ref{app:affective_update}.

\subsection{Runtime and precision boundary}

The partner-local POMDPs supply ordinary active-inference belief updating and policy evidence for each partner. They do not contain \(\beta_k\) as a hidden state. Instead, \(\beta_k\) is maintained as an auxiliary partner-local precision tracker, updated from partner-response likelihood surprisal and then mapped to policy precision \(\gamma_k\). This design keeps the POMDP state space compact and makes the precision pathway experimentally isolable. The cost is that the focal agent cannot form prior expectations over its own future confidence states. Appendix~\ref{app:affective_update} specifies the auxiliary update and the cross-partner policy-selection procedure.
