\begin{abstract}
\emergencystretch=2em
Affect---the fundamental experiential substrate of emotion in psychology and neuroscience---does not merely colour inner experience. It regulates how decisively beliefs become action, mediating the transition from what an agent knows to what it commits to. The active inference framework has formalised this as a metacognitive signal: a running estimate of generative model fitness that modulates action confidence through an inferred precision parameter \cite{Hesp2021DeeplyFeltAffect}. But this formalisation has a structural gap at the multi-agent level. Emotional confidence is inherently relational---we carry it relationship by relationship, earned through each partner's specific behavioural history. We extend the affective precision mechanism to the partner level, giving each social relationship its own model-fitness estimate and independent action-confidence budget, updated from that partner's behavioural evidence alone.

Simulations in a repeated multi-partner trust game reveal dynamics that are behaviourally and psychologically recognisable. Partner-local affective precision tracks partner \emph{predictability} rather than partner \emph{value}: a reliably hostile partner and a reliably cooperative one can both earn high precision because what they share is predictability, not valence. Behavioural gains arise specifically through action sharpening rather than improved belief quality, confirmed by a tracked-only lesion that severs the precision-to-action link while leaving partner-belief updating intact. Engagement reorganises around reliability before payoff differences accumulate. Under abrupt betrayal the mechanism fails in a specific and interpretable way---accumulated social confidence sharpens action around a model that has become stale---revealing the temporal assumption on which adaptive social confidence depends.

The parameters governing these dynamics correspond to psychologically recognisable individual differences in social trust calibration. Varying the precision-gain $\alpha$ and the prior over model fitness produces four behavioural phenotypes---anxious-reactive, hypervigilant, naive-stubborn, and avoidant-rigid---that differ in exploitation sensitivity, betrayal recovery, forgiveness speed, and mixed-volatility discrimination. These signatures map onto empirically characterised variation in human personality and clinical profiles of social trust and attachment. Affect, in this account, is a metacognitive signal regulating social commitment whose dynamics, failure modes, and individual variation reflect the temporal and relational structure of human social life.
\keywords{Active inference \and Affect \and Social cognition \and Multi-agent systems \and Trust \and Individual differences}
\end{abstract}
