\section{Introduction}

Social life requires two things that are routinely conflated. The first is \emph{knowing}: learning what partners are likely to do, building representations of their dispositions and intentions. The second is \emph{committing}: deciding how decisively to act on what you know. A well-calibrated belief that a partner defects 80\,\% of the time is not the same as knowing how much to bet on it. The inference problem and the commitment problem are separable, and most formal models of social cognition address the first while leaving the second implicit.

In psychology and neuroscience, affect---the fundamental experiential substrate of emotion \cite{Barrett2006EmotionParadox}, the underlying valence and arousal that gives emotional experience its character---is what regulates the commitment problem. Affect does not merely colour experience; it mediates the transition from inference to action, modulating how much weight beliefs carry when choices are made. This is computationally significant, not merely phenomenological: an agent that infers well but commits poorly wastes its social knowledge, while one that commits confidently on fragile beliefs exposes itself to systematic exploitation. The affective signal, in this view, is not about representing the social world---it is about regulating how strongly the agent's representation of the social world drives its behaviour.

The active inference framework \cite{Friston2010FreeEnergy,Friston2017ProcessTheory,Parr2022ActiveInference} provides a formal route from this observation to a computational mechanism. Agents minimise variational free energy to update beliefs and evaluate candidate policies by their expected free energy \cite{Friston2015EpistemicValue,DaCosta2020DiscreteSynthesis}. Action selection is precision-weighted: the same beliefs can produce decisive or tentative behaviour depending on $\gamma$, which controls how sharply the agent commits to its most-preferred policy. Hesp et al.\ \cite{Hesp2021DeeplyFeltAffect} made affect endogenous by modelling the agent as actively inferring a higher-order belief $\beta$ over its expected action precision---the subjective fitness of its current generative model. An affective charge signal $\phi$ drives updates to $\beta$: it rises when the model performs better than a prior baseline and falls when it performs worse. Affect, in this formalisation, is a derivative of model fitness rather than reward history---a distinction that allows the signal to be positive toward a reliably hostile partner (whose actions the model predicts well) while remaining informative about when to act confidently.

This formalisation has a gap at the multi-agent level. Emotional confidence is inherently relational. In daily social life, we do not carry the same felt certainty toward every person we know---we carry it relationship by relationship, earned through each partner's specific behavioural history. A person can have a well-validated model of a longtime colleague, a fragile model of a new acquaintance, and an actively revising model of a partner who has recently surprised them, and these three states of social confidence should not share a single global precision budget. Global precision conflates them: it cannot express that you should act decisively with a well-understood partner and cautiously with an unfamiliar one, even when both beliefs are held simultaneously. The psychology of affect demands localisation; the only question is what such localisation produces.

We extend the affective precision mechanism to the partner level. A focal agent interacting with multiple partners maintains a separate belief state and a separate $\beta_k$ estimate for each partner $k$, both updated from that partner's behavioural evidence alone. This allows action confidence to be modulated independently across the agent's social portfolio---more decisive toward partners whose behaviour the model predicts well, more cautious toward those who remain uncertain. The interest of this extension lies not in the factorisation itself, which is forced by the relational structure of affect, but in the dynamics it produces and what those dynamics reveal about the computational basis of social behaviour.

Using a repeated multi-partner trust game with varied decision regimes, we test whether the emergent behavioural dynamics of partner-local affective precision are psychologically recognisable while keeping the current evidence tier explicit. The mechanism is designed to track predictability rather than value. Current post-fix diagnostic evidence supports the expected metacognitive pathway: the same partner models, deployed with different action confidence, alter policy entropy and allocation before payoff is reliable. The corrected H1 readout supports the surprise-over-reward dissociation at smoke scale but requires confirmation and controlled reward-matched diagnostics before carrying the model-fitness claim. Under abrupt betrayal, the corrected selector makes H5 the strongest behavioral confirmation target, revealing the temporal conditions on which adaptive social confidence depends.

Going further, systematic manipulation of the precision-gain parameter $\alpha$ and the prior over model fitness defines four computational phenotype targets---anxious-reactive, hypervigilant, naive-stubborn, and avoidant-rigid---predicted to differ in exploitation sensitivity, betrayal recovery, forgiveness speed, and mixed-volatility discrimination. These targets are intended to connect the computational architecture to individual differences without treating the simulations as clinical validation. The phenotype predictions are consistent with the empirical literature on individual differences in social learning rates \cite{Behrens2008SocialLearning} and volatility tracking \cite{Mathys2011HGF}, and with computational accounts of trust dysregulation in affective and personality disorders \cite{Adams2013ComputationalPsychosis}.

We address three questions. First, does partner-local affective precision track model reliability or realised payoff, and where does its behavioural effect originate---in belief quality or action confidence? Second, what determines the boundary between adaptive and maladaptive precision modulation, and how does this boundary interact with the speed of social change? Third, how do individual differences in the mechanism's parameters shape behavioural phenotypes across social contexts, and what do these phenotypes reveal about the computational basis of human variation in social trust?
