\section{Discussion}
\label{sec:discussion}

\paragraph{Affect as social metacognition.}
Partner-local affective precision is best understood as a form of social metacognition: the agent's estimate not of what a partner will do, but of how well its current model of that partner justifies confident action. The signal is turned inward, toward the agent's own inference quality, rather than outward toward the partner's mental states. This distinguishes the mechanism from theory-of-mind inference, which models what partners believe, know, or intend---a second-order model of another agent's mental states \cite{FrithFrith2012,Pitliya2025ToMActiveInference}. Affective precision does not do this. It tracks the focal agent's confidence in its own model of a partner's \emph{behaviour}, not representations of that partner's mental states. The signal is metacognitive rather than mentalistic.

The two mechanisms are complementary. An agent with sophisticated ToM inference could still be poorly calibrated about when to act confidently on those inferences, particularly if the partner is reliable in some contexts and unreliable in others. Partner-local affective precision addresses exactly this gap: it allows action confidence to track the recent track record of each partner model, independently of the model's representational content \cite{RuizSerra2025Factorised}. The combination would permit an agent to both model partner intentions and maintain a running estimate of how much those models currently justify decisive action---a richer form of social metacognition that may be necessary for human-like performance in complex multi-partner environments.

\paragraph{Individual differences and the computational basis of social trust variation.}
The phenotype programme in Section~\ref{sec:phenotypes} defines a computational target connecting affective precision dynamics to a broader literature on individual differences in social cognition. The key hypothesis is that variation in $\alpha$ and prior initial conditions should not produce merely quantitative differences in outcomes, but qualitatively distinct behavioural signatures that are recognisable as social-trust calibration profiles.

The anxious-reactive profile (high $\alpha$, naive prior) parallels the computational signature of anxious hypervigilance in social contexts, in which threat-sensitive prediction error weighting produces rapid, high-variance oscillations in social confidence \cite{Adams2013ComputationalPsychosis}. The hypervigilant profile (high $\alpha$, cautious prior) parallels paranoid ideation---chronic wariness that resists updating even under sustained positive evidence. Computational psychiatry accounts of psychosis and social paranoia have proposed exactly this mechanism: aberrant precision weighting in the action-confidence pathway produces behaviour that prioritises threat detection at the cost of evidence integration. The avoidant-rigid profile (low $\alpha$, cautious prior) corresponds to avoidant attachment \cite{Cittern2018AttachmentActiveInference}---not hypersensitive to negative signals but chronically unable to let positive evidence accumulate into confident engagement, producing social withdrawal maintained not by fear but by slow updating. The naive-stubborn profile captures a different failure mode: not paranoid but exploitable, failing to sufficiently update social confidence after negative evidence because the updating mechanism itself is sluggish.

Crucially, these phenotypes are not simply high-trust versus low-trust distinctions. They are predicted to differ in the \emph{asymmetry} and \emph{speed} of trust revision, not just its baseline level. The forgiveness paradigm tests this distinction directly by asking whether $\alpha$ and prior produce separable effects on recovery dynamics corresponding to the empirical distinction between forgiveness (willingness to re-engage) and trust repair (restoration of confidence) \cite{Yuan2017ApologyTrustRepair}. The target dissociation is that high-$\alpha$ agents re-approach quickly, whereas cautious-prior agents fail to repair confidence fully even after returning. This distinction is lost in any framework that treats social trust as a scalar quantity.

The mixed-volatility paradigm adds a further dimension: it tests whether individual differences in social confidence calibration translate into a sensitivity-specificity tradeoff in social environments that contain both stable and changing relationships simultaneously. The target prediction is that high-$\alpha$ agents detect genuine partner shifts faster but misattribute noise as signal at higher rates. This would parallel the behavioural signature of social anxiety in trait personality research---enhanced vigilance to threat-relevant social cues at the cost of stable engagement with reliable partners \cite{Behrens2008SocialLearning}. The present framework positions these phenotypes as computational predictions for empirical individual-difference research: fitting $\alpha$ and prior parameters to individual learning curves in iterated trust games, then testing whether estimated parameters predict personality measures and clinical scores, would allow the model to be evaluated against human behavioural data.

\paragraph{The lesion result supports a deployment claim.}
The tracked-only lesion supports the deployment interpretation at the current diagnostic scale: removing the $\beta_k \to \gamma_k$ link makes tracked-only runs match no-affect controls, even though the beta state is still updated. The current evidence therefore supports the claim that partner-local precision changes policy commitment through the action-confidence pathway. The lesion design is the methodological core of this claim because it targets the alternative that $\beta_k$ tracking indirectly biases belief quality, which would have implicated the inference pathway rather than the commitment pathway.

\paragraph{The betrayal result marks the temporal-dependency target.}
The corrected agent-choice selector reveals a specific candidate pattern in the abrupt betrayal task: local affect outperforms no-affect and tracked-only controls on smoke-scale payoff ($1322.3$ versus $1225.0$) while simultaneously showing lower joint accuracy ($0.319$ versus $0.425$). The payoff gain alongside the accuracy reduction suggests that the mechanism is not working through better belief updating. Instead, accumulated $\beta_k$ evidence from the pre-betrayal cooperative phase changes action commitment around a partner model whose reliability has shifted. The partner-specific precision signal may then allow the agent to redirect engagement toward reliable alternatives once cross-partner comparison reveals the divergence, but this remains the priority H5 confirmation target rather than a settled publication-grade claim.

The implication is structural: any confidence mechanism that integrates evidence without a separate volatility channel will exhibit this temporal dependency. The $\beta_k$ tracker's failure mode under abrupt social change is not an implementation bug but an inherent property of social confidence built on evidence accumulation without change detection. This is structurally similar to the challenge faced by hierarchical volatility models \cite{Mathys2011HGF,Behrens2008SocialLearning}, but applied to action confidence rather than learning rates---a level of the hierarchy that existing models do not separately address.

\paragraph{Limitations.}
Four design features bound the current claims. First, $\beta_k$ is an external auxiliary tracker; in the full Hesp et al.\ hierarchy it is a proper inferred hidden state, allowing the agent to form prior expectations over its own future affective states and to plan around anticipated changes in partner reliability. Making $\beta_k$ a variational variable would enable richer prospective regulation of action confidence. Second, the current implementation keeps $q(\beta_k)$ categorical rather than continuous. This preserves the inverse-beta precision convention and keeps the update identifiable, but it discretises the precision dynamics. Third, partners are scripted rather than themselves being active-inference agents; the mechanism is tested in isolation from the reciprocal dynamics that characterise real social interaction. Fourth, the phenotype taxonomy in Section~\ref{sec:phenotypes} is currently predictive rather than validated: the experiments test whether different $\alpha$ and prior values produce qualitatively distinct behavioural profiles, and the manuscript connects those targets to human individual differences in social trust. A stronger claim---that individual differences in human social trust calibration arise specifically from variation in an affective precision mechanism with this structure---would require fitting the model to human behavioural data and establishing that parameter estimates predict theoretically relevant outcomes. The phenotype predictions are offered as a scaffold for such empirical work, not as an established correspondence.

\paragraph{Future directions.}
The most important near-term extension is empirical: fitting the affective precision model to human behavioural data from iterated trust game paradigms with individual difference measures. Sections~\ref{sec:phenotypes_quadrants}--\ref{sec:phenotypes_mixed} generate specific, quantitative predictions about exploitation sensitivity, betrayal recovery time, forgiveness speed, and mixed-volatility discrimination that differ across phenotype quadrants and are measurable from behavioural time series without neuroimaging. The $\alpha$ and prior parameters could be estimated per participant and tested against personality measures (Big Five dimensions, attachment style) and clinical scores (social anxiety, paranoid ideation) to assess whether the computational phenotypes correspond to real psychological variation. Such a study would also allow a within-design test of whether global versus partner-local precision provides better fits to individual behavioural trajectories.

A second priority is the global-$\beta$ ablation: whether partner-local precision is necessary, or whether a single shared precision estimate updated from aggregate social experience produces comparable stable-regime gains. The mixed-volatility environment in Section~\ref{sec:phenotypes_mixed} constitutes a direct test of this question. The predicted outcome is that a shared tracker would over-suppress confidence toward reliably predictable partners or under-suppress it toward uncertain ones, depending on social environment composition, and crucially that it would fail to produce the target discrimination pattern in the heterogeneous-volatility condition.

Third, the mixed-volatility environment points toward richer ecological settings: social environments with communication, reputation signalling, and coalition structure would introduce new sources of reliability evidence that the affective signal could in principle track. Extending partner-local $\beta_k$ dynamics to incorporate evidence from reputation signals and third-party observations is a natural next step that connects to the indirect-reciprocity literature.

Finally, combining partner-local affective precision with ToM-style belief inference \cite{Pitliya2025ToMActiveInference} would allow future work to study whether metacognitive action-confidence signals and mentalistic partner-state models interact during social decision-making. An agent with both mechanisms could be confident in its ToM inferences for some partners and uncertain for others---a richer form of social metacognition that may be necessary for human-like performance in complex multi-partner environments. Whether metacognitive confidence and mentalistic inference are functionally coupled or independent in human social cognition is an open empirical question that the present framework directly motivates.

\section{Conclusion}

Predicting what a partner will do and deciding how confidently to act on that prediction are separable problems. Partner-local affective precision addresses the second. By giving each social relationship its own model-fitness estimate, the mechanism allows action confidence to be calibrated independently across a social portfolio---more decisive toward partners whose behaviour the model predicts well, more cautious toward those who remain uncertain. Current diagnostic evidence supports reliability tracking, metacognitive rather than epistemic deployment, and a volatility boundary condition that reveals the temporal assumption on which social confidence depends. The parameters governing these dynamics---precision gain and prior over model fitness---define a family of behavioural phenotype targets for individual-difference research, positioning the framework as a mechanistically grounded and empirically tractable account of social affect.
