\section{Experiments and Analysis}

\subsection{Experiment cards}

H0 tests whether policy-space openness gates observable affect effects. If the policy posterior is already saturated, beta can move without changing behavior. Open graded action spaces, partner-choice regimes, and volatility windows should leave more policies alive and therefore expose precision effects.

H1 tests whether partner-local precision tracks predictive reliability rather than realized reward. The critical dissociation is a predictable low-reward partner: a reward cache should be low, while a model-fitness signal can be high if the agent can predict the partner.

H2 tests deployment by comparing affective agents with tracked-only or no-affect controls. The expected pattern is that belief readouts remain similar while policy entropy, payoff, or partner choice diverges in open regimes.

H3 tests stress. Abrupt stance switches, betrayal, and volatility should amplify the precision channel. The desired result is not necessarily recovery success. A valid boundary-condition result is precision-driven misdeployment: lowered entropy, altered returns or reallocation, and worse outcomes after a sudden shift.

H4 tests social choice. Partner-specific precision should alter whom the agent selects, avoids, probes, or returns to, even when total payoff is still flat.

H5 tests whether clinical-like precision perturbations separate in beta dynamics and behavior. H5 is supplemental because current runs use five seeds per variant and because no human or clinical validation data are included.

\subsection{Evidence base}

\begin{table}[t]
\centering
\caption{Audited run sizes included in this draft.}
\label{tab:run-sizes}
\begin{tabularx}{\linewidth}{lXccc}
\toprule
Experiment & Result role & Variants & Seeds & Rounds \\
\midrule
H0 graded choice & Supporting openness evidence & 2 & 5 each & 200 \\
H2 lesion open & Supporting deployment evidence & 3 & 5 each & 200 \\
H4 partner choice & Supporting social-choice evidence & 2 & 5 each & 200 \\
H1 confirmation & Primary model-fitness evidence & 2 & 30 each & 200 \\
H3 confirmation & Primary betrayal boundary evidence & 3 & 30 each & 120 \\
H3 abrupt sensitivity & Primary shock-regime follow-up & 12 & 30 each & 120 \\
H3 gradual sensitivity & Primary shock-regime follow-up & 8 & 30 each & 120 \\
H5 clinical dynamics & Supplemental perturbation dynamics & 4 & 5 each & 200 \\
H5 clinical betrayal & Supplemental perturbation dynamics & 4 & 5 each & 120 \\
H5 affect sensitivity & Supplemental perturbation dynamics & 6 & 5 each & 120 \\
\bottomrule
\end{tabularx}
\end{table}

The primary measures are total payoff, mean policy entropy, type/stance/joint belief accuracy, predictive log score, partner-selection entropy, partner-selection rates, beta statistics, precision-surprise and precision-payoff associations, reencounter behavior after betrayal, payoff conditional on return, and low-entropy misdeployment rates. Analysis outputs report mean differences, bootstrap confidence intervals for selected effects, and pairwise tests for payoff comparisons. This draft treats those statistics as current evidence but not as a final preregistered analysis package.
