\appendix

\section{Claim and Non-Claim Checklist}

\begin{longtable}{p{0.30\linewidth}p{0.32\linewidth}p{0.30\linewidth}}
\caption{Draft claim discipline.}\label{tab:claim-checklist}\\
\toprule
Claim & Current support & How to write it \\
\midrule
\endfirsthead
\toprule
Claim & Current support & How to write it \\
\midrule
\endhead
Policy openness gates affect effects & H0/H2 five-seed open-regime evidence; saturated regimes weaker & ``Policy openness gates visible affective deployment.'' \\
Partner-local precision tracks model fitness more than reward & H1 30-seed confirmation; precision-surprise association exceeds precision-payoff association & ``Precision is predictive reliability, not cached partner value.'' \\
Affect changes deployment more than belief inference & H2 five-seed lesion; belief readouts similar while entropy/payoff move & ``Knowing-without-using dissociation.'' \\
Stress amplifies the channel & H3 30-seed betrayal evidence; entropy lower and behavior changes & ``Stress exposes both mechanism and failure mode.'' \\
Affect improves recovery after betrayal & Not supported in current H3 confirmation & Do not claim. Say abrupt betrayal exposes misdeployment risk. \\
Generic caution fixes H3 & Not supported; cautious variants often hurt payoff & Say shock shape matters more than generic caution. \\
Partner-specific precision guides social choice & H4 five-seed partner-choice shift with flat payoff & ``Partner choice is the readout; payoff is secondary here.'' \\
Clinical variants are validated phenotypes & Not supported & Say ``clinical-like parameter perturbations,'' not diagnoses. \\
Partner-local beta beats global beta & Not yet tested & List as future global-beta ablation. \\
\bottomrule
\end{longtable}

\section{Figure Plan}

\begin{enumerate}
  \item \textbf{Figure 1: Model schematic.} Show partner action prediction error $\rightarrow$ partner-specific $\beta_k$ / precision$_k$ $\rightarrow$ partner-local $\gamma_k$ $\rightarrow$ policy posterior shift $\rightarrow$ action/partner choice $\rightarrow$ payoff, reallocation, or misdeployment. Include official partner-local \texttt{pymdp.Agent} instances, hidden factors, observations, external beta tracker, beta-to-gamma modulation, and tracked-only lesion.
  \item \textbf{Figure 2: Open-regime deployment.} Show total payoff, policy entropy, and belief readouts for affect versus no-affect/lesion in the open graded regime.
  \item \textbf{Figure 3: Model fitness versus reward.} Show precision-surprise and precision-payoff associations, reliability-over-reward effect, and no payoff advantage.
  \item \textbf{Figure 4: Betrayal boundary condition.} Show payoff, entropy, reencounters, payoff-on-return, and wrong-type rate.
  \item \textbf{Figure 5: Shock shape and sensitivity.} Compare abrupt and gradual betrayal, including default affect, no-affect/lesion baseline, and cautious variants.
  \item \textbf{Supplement: H4 social choice and H5 perturbation dynamics.} Show partner-selection rates, partner-selection entropy, beta range, beta autocorrelation, action-flip rate, and selection entropy.
\end{enumerate}

\section{Result Summary Table}

\begin{longtable}{p{0.15\linewidth}p{0.48\linewidth}p{0.30\linewidth}}
\caption{Numerical result summary for draft writing.}\label{tab:result-summary}\\
\toprule
Result & Primary numbers & Interpretation \\
\midrule
\endfirsthead
\toprule
Result & Primary numbers & Interpretation \\
\midrule
\endhead
R1 H0/H2 openness & Five seeds: affect payoff $1884.6$, entropy $7.94$; baseline payoff $1859.4$, entropy $8.80$ & Open graded policy spaces let precision move deployment. \\
R2 H1 model fitness & 30 seeds: payoff $534.6$ vs $542.1$; $|r_{\mathrm{precision},\mathrm{surprise}}|=0.701$ vs $|r_{\mathrm{precision},\mathrm{payoff}}|=0.419$; reliability-over-reward $+0.096$ CI $[0.027,0.164]$ & Precision tracks predictive reliability more than reward. \\
R3 H2 lesion & Five seeds: affect joint accuracy $0.336$, stance accuracy $0.816$, entropy $7.94$; lesion joint accuracy $0.339$, stance accuracy $0.855$, entropy $8.80$ & Beliefs similar; deployment differs. \\
R4 H4 social choice & Five seeds: payoff $393.6$ vs $393.2$; entropy $3.99$ vs $4.83$; selection rates move from $[0.348,0.318,0.162,0.172]$ to $[0.366,0.316,0.138,0.180]$ & Choice distribution changes before payoff. \\
R5 H3 betrayal & 30 seeds: payoff $1136.1$ vs $1172.1$; CI $[-63.65,-10.93]$; entropy $8.38$ vs $8.74$; reencounters $4.4$ vs $6.1$; wrong-type rate $0.237$ vs $0.165$ & Abrupt betrayal exposes misdeployment. \\
R6 H3 shock shape & 30 seeds: abrupt baseline $1153.6$, default affect $1140.4$, combined caution $1115.0$; gradual baseline $1148.9$, default affect $1147.6$, combined caution $1118.3$ & Gradual shock removes default affect payoff penalty; generic caution does not rescue payoff. \\
R7 H5 perturbations & Five seeds: dynamics beta range borderline $1.318$, depression $1.435$, affect $0.950$, alexithymia $0.149$ & Perturbations separate dynamics, not clinical validity. \\
\bottomrule
\end{longtable}
